Path Analysis Data Development - Chapter 2

### 2016-02-24 sem__data_development.Rmd
Update using corrected stems data.
I used the corrected stems data developed in the 'corrected-dbh.R' script, which should be more accurate (i.e. have more corrections). I also removed the decomposition of topographic wetness index to its own script.

## Load data
```{r pairs helper functions, echo=FALSE}
panel.hist <- function(x, ...)
{
      usr <- par("usr"); on.exit(par(usr))
      par(usr = c(usr[1:2], 0, 1.5) )
      h <- hist(x, plot = FALSE)
      breaks <- h$breaks; nB <- length(breaks)
      y <- h$counts; y <- y/max(y)
      rect(breaks[-nB], 0, breaks[-1], y, col = "cyan", ...)
}
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
      usr <- par("usr"); on.exit(par(usr))
      par(usr = c(0, 1, 0, 1))
      r <- abs(cor(x, y, use = "na.or.complete"))
      txt <- format(c(r, 0.123456789), digits = digits)[1]
      txt <- paste0(prefix, txt)
      if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
      text(0.5, 0.5, txt, cex = cex.cor * r)
}
```
```{r load ecological data, echo=FALSE}
# Ecological data: leaf counts, infection
load("~/GitHub/superspreaders/stems_plots.RData")

@@

write.csv(plots_env, "analysis/data/env_gis_variables.csv", row.names = F)
plots_env <- read.csv("analysis/data/env_gis_variables.csv")
```

@@

# BUSH01 plot included. I think we have limited ecological data for this plot, and no microclimate data. So, I am going to exclude it from all the data sets
# plots_env <- filter(plots_env, plotid != "BUSH01")
stems <- as.tbl(stems)
stems
stems <- select(stems, -X) 
stems <- rename(stems, plotid = plot) %>% filter(plotid != "BUSH01")

@@

### Save data with BUSH01 excluded
```{r save data}
write.csv(plots_env, "analysis/data/env_gis_variables.csv", row.names = F)
write.csv(rain_df, "analysis/data/rain_data_long.csv", row.names = F)
write.csv(rs_temps, "analysis/data/rs_temps_data.csv", row.names = F)
write.csv(summer_temps, "analysis/data/summer_temps_data.csv", row.names = F)
write.csv(stems, "analysis/data/stems_noBUSH01.csv", row.names = F)
rm(plots_temps); rm(plots_rain)
```

## Starting Point with Revised Data: Reload data that has BUSH01 excluded
Still need to load the .RData source at the beginning of this document, line 32.
```{r reload data}
plots_env <- read.csv("analysis/data/env_gis_variables.csv")
rain_df <- read.csv("analysis/data/rain_data_long.csv")
rs_temps <- read.csv("analysis/data/rs_temps_data.csv")
summer_temps <- read.csv("analysis/data/summer_temps_data.csv")
stems <- read.csv("analysis/data/stems_noBUSH01.csv")
```

## Plot Contributing Area vs. Slope
This is a suggestion from Ryan to plot the contributing area vs. the slope for each of the plots in order to tease out how either influences infection and hosts. Decomposing TWI to determine which factor has a greater influence on disease dynamics. 
```{r decomposing TWI influence}
library(ggplot2)
# par(mfrow = c(3,1))
names(plots_env)
summary(plots_env)
## 'facc_ws15m' is the estimate from the 'r.watershed' function in GRASS GIS

# Plot estimates from two different functions agains each other.
qplot(flowacc15m, facc_ws15m, data = plots_env) #+
      geom_text(aes(label = plotid))
cor(plots_env$flowacc15m, plots_env$facc_ws15m)
## There are 9 plots that really stand out from the others.
## The r.watershed calculation provides similar, but generally smaller values.
qplot(flowacc15m, facc_ws15m, data = filter(plots_env, flowacc15m > 300),
      main = "'High' Flow Accumulation Value Plots") +
      geom_text(aes(label = plotid))
qplot(flowacc15m, facc_ws15m, data = filter(plots_env, flowacc15m < 300),
      main = "'Low' Flow Accumulation Value Plots") +
      geom_text(aes(label = plotid))
## Much of the variability between the two estimates stems from a few plots.

# Plot flow accumulation vs. slope
qplot(slope15m, flowacc15m, data = plots_env) #+
      geom_text(aes(label = plotid))
## Can't see much pattern (if there is any) because of a few big values
      cor(plots_env$slope15m, plots_env$flowacc15m)
      ## negative, but weak

qplot(slope15m, flowacc15m, data = filter(plots_env, flowacc15m < 300), 
      main = "Flow Accumulation <300") #+
      geom_text(aes(label = plotid))
## Generally, a shallower slope means higher flow accumulation, but it's not a strong pattern

qplot(slope15m, flowacc15m, data = filter(plots_env, flowacc15m > 300), 
      main = "Flow Accumulation >300 ") #+
      geom_text(aes(label = plotid))
## The set of high flow accumulation plots shows an opposite general trend in the relationship between flow accumulation and slope. It's a small sample size, which leads me to consider the general trend from the large sample size being closer to the real relationship.

qplot(slope15m, facc_ws15m, data = plots_env)
cor(plots_env$slope15m, plots_env$facc_ws15m)
## weakly negative correlation between slope and flow accumulation

qplot(slope15m, facc_ws15m, data = filter(plots_env, facc_ws15m < 300), 
      main = "Flow Accumulation (r.watershed) <300") #+
      geom_text(aes(label = plotid))
qplot(slope15m, facc_ws15m, data = filter(plots_env, facc_ws15m > 300), 
      main = "Flow Accumulation (r.watershed) >300") #+
      geom_text(aes(label = plotid))

qplot(slope15m, facc_ws15m, data = filter(plots_env, facc_ws15m > 50 & facc_ws15m < 200), 
      main = "facc 50 - 200") +
      geom_text(aes(label = plotid))
qplot(slope15m, facc_ws15m, data = filter(plots_env, facc_ws15m > 200 & facc_ws15m < 1000), 
      main = "facc 200 - 1000") +
      geom_text(aes(label = plotid))
qplot(slope15m, facc_ws15m, data = filter(plots_env, facc_ws15m > 1000), 
      main = "facc >1000") +
      geom_text(aes(label = plotid))

# pairs(select(plots_env, -plotid, -starts_with("rain")), 
#       lower.panel = panel.cor, diag.panel = panel.hist)
# pairs(select(plots_env, -plotid, -starts_with("rain")) %>% 
#             filter(flowacc15m < 5), 
#       lower.panel = panel.cor, diag.panel = panel.hist,
#       main = "Flow accumualtion < 5")
# pairs(select(plots_env, -plotid, -starts_with("rain")) %>% 
#             filter(flowacc15m > 100 & flowacc15m < 1000), 
#       lower.panel = panel.cor, diag.panel = panel.hist,
#       main = "Flow accumulation 100-1000")
# pairs(select(plots_env, -plotid, -starts_with("rain")) %>% 
#             filter(flowacc15m > 1000), 
#       lower.panel = panel.cor, diag.panel = panel.hist,
#       main = "Flow accumulation > 1000")

plot(flowacc15m ~ slope15m, data = filter(plots_env, flowacc15m > 300))
pairs(select(plots_env, flowacc15m, slope15m) %>% filter(flowacc15m > 300), 
      lower.panel = panel.cor, diag.panel = panel.hist)

library(manipulate)
manipulate(
      qplot(slope15m, flowacc15m, data = plots_env,
           axes = axes,
           cex = cex
           #pch = if(pch) 19 else 1
           ),
      axes = checkbox(TRUE, "Show axes"),
      cex = slider(0, 5, initial = 1, step = 0.1, label = "Point size"),
      pch = button("Fill points")
)

dataplot <- function(dat){
      name <- sys.call()[[2]]
      vars <- as.list(names(dat))
      e <- new.env()
      e$data <- name
      manipulate(
            {
                  form = as.formula(paste(y,x, sep = "~"))
                  plot(form, data = dat, main = as.character(name), las = 1)
                  e$form <- form
            },
            x = do.call(picker, c(vars, initial = vars[1])),
            y = do.call(picker, c(vars, initial = vars[2]))
      )
      invisible(e)
}
f <- dataplot(plots_env)
```

## Examine the bay laurel data at individual and plot levels
```{r examine umca data}
summary(plots_umca)
# This only has number of infected, uninfected, and total, which must apply to living stems only. I want to analyze leaf counts.

unique(plots_umca$plot)# 185 plots with bay laurel
umca_plots <- filter(plots_umca, plot != "BUSH01")# now 184 plots
rm(plots_umca)
summary(bay_laurel) # This has stem-level data on bay laurel.
unique(bay_laurel$plot)
bay_laurel <- filter(bay_laurel, plot != "BUSH01")
```

I can use the stem-level data to calculate plot level leaf count variables (sum, average) for each plot and year.
```{r calculate plot level leaf counts}
live_umca <- bay_laurel %>% filter(status == "Alive")
filter(live_umca, is.na(slc))
# Spot checking some of these shows that NAs are reasonable
umca_plots <- left_join(
      umca_plots, bay_laurel %>% filter(status == "Alive") %>% 
            group_by(plot, year) %>% 
            summarise(avg_lfct = mean(slc, na.rm = T),
                      tot_lfct = sum(slc, na.rm = T)),
      by = c("plot", "year"))
summary(umca_plots)
```

Explore the NAs in the UMCA leaf counts.
```{r examine lfct NAs}
filter(umca_plots, is.na(avg_lfct))
filter(bay_laurel, plot == "YAHNG02", status == "Alive")
# I'm not sure why this wasn't excluded from the calculations with the na.rm = T option
# I'm removing it manually from the `umca_plots` data frame
umca_plots <- umca_plots %>% filter(avg_lfct != "NaN")
summary(umca_plots)
```

So now I have a plot-level data frame of UMCA infection for each year in terms of symptomatic leaf counts and number of un/infected stems. I want to join this with some of the plot-level environmental data.
```{r join env and umca data}
summary(plots_env)

@@

# Join the environmental data to the bay laurel infection data
umca_plots <- left_join(select(umca_plots, -species), plots_env, by = c("plot" = "plotid"))
summary(umca_plots)
filter(umca_plots, is.na(lon_utm))
# NAs due to no topo data for abandoned SUGAR02; remove SUGAR02
umca_plots <- na.omit(umca_plots)
umca_plots <- rename(umca_plots, plotid = plot, sample_year = year)
umca_plots$plotid <- tolower(umca_plots$plotid)

@@

```{r join umca and qusp data}
umca_plots
umca_plots$sample_year <- as.integer(umca_plots$sample_year)
plot_qusp # right now this is broken down by oak species
qusp_plots <- plot_qusp
rm(plot_qusp)
summary(qusp_plots)

oak_sod <- qusp_plots %>% group_by(plot, year) %>%
      summarise(uninf_oak_ct = sum(uninfected_oak_ct), 
                inf_oak_ct = sum(infected_oak_ct), 
                tot_oak_ct = sum(c(uninfected_oak_ct, infected_oak_ct)))
oak_sod
oak_sod <- rename(oak_sod, plotid = plot, sample_year = year)
oak_sod$plotid <- tolower(oak_sod$plotid)

unique(umca_plots$plotid) %in% unique(oak_sod$plotid)

@@

os.us.diversity
oak_sod <- left_join(oak_sod, os.us.diversity, by = c("plotid"))
summary(oak_sod)
filter(oak_sod, is.na(elev_m))
# NAs are plots that were abandoned early in the study
oak_sod <- filter(oak_sod, plotid != "duer01", plotid != "ponti01", plotid != "sweet01")

summary(filter(oak_sod, is.na(avg_tmax_rs)))
# Many of the NAs are due to the 2003 sample year
oak_sod <- filter(oak_sod, sample_year != 2003)
filter(oak_sod, is.na(avg_tmax_rs))
# BUSH01 plot; I'm dropping that from the data set
oak_sod <- filter(oak_sod, plotid != "bush01")
summary(oak_sod)

@@

filter(oak_sod, sample_year == 2004) # 88 records
# 2005 appears to be the first full year of observing all the plots
filter(oak_sod, sample_year == 2005) # 170 records
filter(oak_sod, sample_year == 2006) # 172 records
filter(oak_sod, sample_year == 2007) # 171 records
filter(oak_sod, sample_year == 2008) # 171 records
filter(oak_sod, sample_year == 2009) # 170 records
filter(oak_sod, sample_year == 2010) # 170 records
filter(oak_sod, sample_year == 2011) # 170 records; final year for Sarah's 2nd chapter
filter(oak_sod, sample_year == 2012) # 169 records
filter(oak_sod, sample_year == 2014) # 168 records
# In Sarah's analysis we end up using only 163 plots
```

@@

```{r oak_sod pairs plot}
class(oak_sod)
oak_sod <- ungroup(oak_sod)
summary(oak_sod)
pairs(select(oak_sod, elev_m, starts_with("hrs"), starts_with("avg_t"), contains("rain_tot")), lower.panel = panel.cor, diag.panel = panel.hist, main = "Oak-SOD Rainfall & Temperature Data")

@@

### Save path analysis data environment
```{r save data environment}
save.image("~/GitHub/superspreaders/pathmodel_data_20160112.RData")
```

>>>>

## Load data
```{r pairs helper functions, echo=FALSE}
panel.hist <- function(x, ...)
{
      usr <- par("usr"); on.exit(par(usr))
      par(usr = c(usr[1:2], 0, 1.5) )
      h <- hist(x, plot = FALSE)
      breaks <- h$breaks; nB <- length(breaks)
      y <- h$counts; y <- y/max(y)
      rect(breaks[-nB], 0, breaks[-1], y, col = "cyan", ...)
}
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
      usr <- par("usr"); on.exit(par(usr))
      par(usr = c(0, 1, 0, 1))
      r <- abs(cor(x, y, use = "na.or.complete"))
      txt <- format(c(r, 0.123456789), digits = digits)[1]
      txt <- paste0(prefix, txt)
      if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
      text(0.5, 0.5, txt, cex = cex.cor * r)
}
```
```{r load ecological data, echo=FALSE}
# Ecological data: leaf counts, infection
load("~/GitHub/superspreaders/stems_plots.RData")
stems <- read.csv("analysis/data/tagged-stems-corrected.csv")

@@

write.csv(plots_env, "analysis/data/env_gis_variables.csv", row.names = F)
# plots_env <- read.csv("analysis/data/env_gis_variables.csv")
```

@@

# BUSH01 plot included. I think we have limited ecological data for this plot, and no microclimate data. So, I am going to exclude it from all the data sets
# plots_env <- filter(plots_env, plotid != "BUSH01")
stems <- as.tbl(stems)
stems
# stems <- select(stems, -X) 
unique(filter(stems, plot_status == "Abandoned")$plotid)
stems <- stems %>% filter(plotid != "bush01", plotid != "ponti01", 
                                plotid != "sweet01")
stems$plotid <- droplevels(stems$plotid)

@@

### Save data with BUSH01 excluded
```{r save data}
write.csv(plots_env, "analysis/data/env_gis_variables.csv", row.names = F)
write.csv(rain_df, "analysis/data/rain_data_long.csv", row.names = F)
write.csv(rs_temps, "analysis/data/rs_temps_data.csv", row.names = F)
write.csv(summer_temps, "analysis/data/summer_temps_data.csv", row.names = F)
# write.csv(stems, "analysis/data/stems_noBUSH01.csv", row.names = F)
# rm(plots_temps); rm(plots_rain)
```

## Starting Point with Revised Data: Reload data that has BUSH01 excluded
Still need to load the .RData source at the beginning of this document, line 32.
```{r reload data}
plots_env <- read.csv("analysis/data/env_gis_variables.csv")
rain_df <- read.csv("analysis/data/rain_data_long.csv")
rs_temps <- read.csv("analysis/data/rs_temps_data.csv")
summer_temps <- read.csv("analysis/data/summer_temps_data.csv")
# stems <- read.csv("analysis/data/stems_noBUSH01.csv")
```

## Examine the bay laurel data at individual and plot levels
```{r examine umca data}
bay_laurel <- filter(stems, species == "umca") %>% 
      select(plotid, cluster:stem_status, umca_slc, notes:dbh_entry)
bay_laurel$species <- droplevels(bay_laurel$species)
summary(bay_laurel) # This has stem-level data on bay laurel.
length(unique(bay_laurel$plotid))# 183 plots with bay laurel
# bay_laurel <- filter(bay_laurel, plot != "BUSH01")

# Data frame of UMCA plots only
umca_plots <- bay_laurel %>%
      # select(plotid, date, species, umca_slc, year, stem_status) %>%
      group_by(plotid, year, species) %>%
      filter(species == "umca", stem_status == "Alive") %>%
      summarise(uninfected_bay_ct = length(which(umca_slc == 0)), 
                infected_bay_ct = length(which(umca_slc > 0)), 
                tot_bay = length(species)) 
summary(umca_plots)

# length(unique(umca_plots$plotid))# 183 plots with bay laurel
# umca_plots <- filter(plots_umca, plot != "BUSH01")
rm(plots_umca)
```

I can use the stem-level data to calculate plot level leaf count variables (sum, average) for each plot and year.
```{r calculate plot level leaf counts}
live_umca <- bay_laurel %>% filter(stem_status == "Alive")
filter(live_umca, is.na(umca_slc))
# Spot checking some of these shows that NAs are reasonable
umca_plots <- left_join(
      umca_plots, 
      live_umca %>% 
            group_by(plotid, year) %>% 
            summarise(avg_lfct = mean(umca_slc, na.rm = T),
                      tot_lfct = sum(umca_slc, na.rm = T)),
      by = c("plotid", "year"))
summary(umca_plots)
```

Explore the NAs in the UMCA leaf counts.
```{r examine lfct NAs}
filter(umca_plots, is.na(avg_lfct))
filter(bay_laurel, plotid == "yahng02", stem_status == "Alive")
# I'm not sure why this wasn't excluded from the calculations with the na.rm = T option
# I'm removing it manually from the `umca_plots` data frame
umca_plots <- umca_plots %>% filter(avg_lfct != "NaN")
summary(umca_plots)
```

So now I have a plot-level data frame of UMCA infection for each year in terms of symptomatic leaf counts and number of un/infected stems. I want to join this with some of the plot-level environmental data.
```{r join env and umca data}
summary(plots_env)
plots_env$plotid <- tolower(plots_env$plotid)

@@

# Join the environmental data to the bay laurel infection data
umca_plots <- left_join(select(umca_plots, -species), plots_env, by = "plotid")
summary(umca_plots)
# filter(umca_plots, is.na(lon_utm))
# NAs due to no topo data for abandoned SUGAR02; remove SUGAR02
# umca_plots <- na.omit(umca_plots)
umca_plots <- rename(umca_plots, sample_year = year)
# umca_plots$plotid <- tolower(umca_plots$plotid)

@@

```{r join umca and qusp data}
umca_plots
umca_plots$sample_year <- as.integer(umca_plots$sample_year)
plot_qusp # right now this is broken down by oak species
qusp4_plots <- stems %>%
      select(plotid, date, species, canker, year, stem_status) %>%
      filter(species == "quag" | species == "quke" | species == "quch" |
                   species == "quwi", stem_status == "Alive") %>% 
      group_by(plotid, year, species) %>% 
      summarise(uninfected_oak_ct = length(which(canker==0)), 
                infected_oak_ct = length(which(canker > 0)), 
                tot_oak = length(species))

# rm(plot_qusp)
summary(qusp4_plots)

oak_sod <- qusp4_plots %>% group_by(plotid, year) %>%
      summarise(uninf_oak_ct = sum(uninfected_oak_ct), 
                inf_oak_ct = sum(infected_oak_ct), 
                tot_oak_ct = sum(c(uninfected_oak_ct, infected_oak_ct)))
oak_sod
oak_sod <- rename(oak_sod, sample_year = year)
# oak_sod$plotid <- tolower(oak_sod$plotid)

# unique(umca_plots$plotid) %in% unique(oak_sod$plotid)

@@

os.us.diversity
oak_sod <- left_join(oak_sod, os.us.diversity, by = c("plotid"))
summary(oak_sod)
filter(oak_sod, is.na(avg_tmax_rs))# NAs are plots sampled in 2003
filter(oak_sod, is.na(H.2014))# NAs are plots lost before 2014
summary(filter(oak_sod, is.na(us.J.2011)))# zero or missing data for understory richness
# oak_sod <- filter(oak_sod, plotid != "duer01", plotid != "ponti01", plotid != "sweet01")

oak_sod <- filter(oak_sod, sample_year != 2003)
filter(oak_sod, is.na(avg_tmax_rs))
# BUSH01 plot; I'm dropping that from the data set
# oak_sod <- filter(oak_sod, plotid != "bush01")
summary(oak_sod)

@@

filter(oak_sod, sample_year == 2004) # 88 records
# 2005 appears to be the first full year of observing all the plots
filter(oak_sod, sample_year == 2005) # 165 records
filter(oak_sod, sample_year == 2006) # 167 records
filter(oak_sod, sample_year == 2007) # 164 records
filter(oak_sod, sample_year == 2008) # 163 records
filter(oak_sod, sample_year == 2009) # 162 records
filter(oak_sod, sample_year == 2010) # 161 records
filter(oak_sod, sample_year == 2011) # 159 records; final year for Sarah's 2nd chapter
filter(oak_sod, sample_year == 2012) # 156 records
filter(oak_sod, sample_year == 2014) # 156 records
# In Sarah's analysis we end up using 163 plots
```

@@

```{r oak_sod pairs plot}
class(oak_sod)
oak_sod <- ungroup(oak_sod)
summary(oak_sod)
# drop rainy days interpolations, they aren't reliable; keep voronoi neighbor method
oak_sod <- select(oak_sod, -rain_days_2d, -rain_days_3d, -rain_days_r)
pairs(select(oak_sod, elev_m, starts_with("hrs"), starts_with("avg_t"), contains("rain_tot")), lower.panel = panel.cor, diag.panel = panel.hist, main = "Oak-SOD Rainfall & Temperature Data")

@@

### Save path analysis data environment
```{r save data environment}
# save.image("~/GitHub/superspreaders/pathmodel_data_20160112.RData")
```

### Update path analysis data
Updating the path analysis data with the temperature during wet days variables.
```{r update data}
# load("pathmodel_data_20160112.RData")
wet_hrs <- read.csv("analysis/data/wet-days-temperature-variables.csv")
save.image("~/GitHub/superspreaders/pathmodel_data.RData")
```
